---
title: "R Notebook"
output: html_notebook
---

### Question 1.
----
#### a)

```{r}

bwgt <- read.csv("Q1.csv", header=T)
bwgt$Race <- as.factor(bwgt$Race)
bwgt$Smoke <- as.factor(bwgt$Smoke)


M0Design <- model.matrix(Bwtkg ~ 1, data=bwgt)
M1Design <- model.matrix(Bwtkg ~ Smoke + Lwtkg, data=bwgt)
M2Design <- model.matrix(Bwtkg ~ Smoke + Lwtkg + C(Race, sum), data=bwgt, x=T)

M0Design

```

```{r}
M1Design
```

```{r}
#options(contrasts = rep ("contr.treatment", 2))  
#constraintBwgt <- bwgt
#attr (bwgt$Race, "contrasts") <- contr.poly

M2Design
```

#### b)
```{r}
M0 <- lm(Bwtkg ~ 1, data=bwgt, x=T)
M1 <- lm(Bwtkg ~ C(Smoke, sum) + Lwtkg, data=bwgt, x=T)
M2 <- lm(Bwtkg ~ C(Smoke, sum) + Lwtkg + C(Race, sum), data=bwgt, x=T)

summary(M0)
```

This is the basic model with just the population mean compared with the Bwtg response variable. We do not have any predictors assigned to this model.


```{r}
summary(M1)
```

p values show that Smoking and Lwtkg are not very significant predictors.
R^2 is quite low, and F-statistic p value also shows low significance


```{r}
summary(M2)
```

This model seems to be the best out of the three as we find two significant predictors:
 - Smoke is significant to 99%
 - Race 1 aka white race is significant to 90%
R^2 is much higher than M1 and F-stat p-value shows significance to reject $H_0$ of group means being equal at the 90% level.

#### c)  

##### &ensp;i.
```{r}
M2.x <- matrix(M2$x, nrow=20)
M2.y <- matrix(bwgt$Bwtkg, nrow=20, ncol=1)

M2.xtx <- t(M2.x) %*% M2.x
M2.inv.xtx <- solve(M2.xtx)

M2.beta <- M2.inv.xtx%*%t(M2.x)%*%M2.y
M2.beta

```

##### &ensp;ii.
```{r}
M2.Hat <- M2.x %*% M2.inv.xtx %*% t(M2.x)

#  P     =          I              -    H
M2.Proj <- diag(nrow=nrow(M2.Hat)) - M2.Hat

# residuals = Py
M2.ehat <- M2.Proj%*%M2.y

# Residual standard error
M2.RSE <- sqrt(sum(M2.ehat^2) / df.residual(M2))
M2.RSE
```


##### &ensp;iii.
```{r}
# MSE = yT*ehat / (n-p)
M2.varhat <- (t(M2.y)%*%M2.ehat)/(length(M2.y)-length(bwgt))
M2.varhat

# SE[Bhat] = sigmahat (or RSE) * diagonal elements of M2.inv.xtx
M2.coefSE <- M2.RSE * sqrt(diag(M2.inv.xtx))
M2.coefSE
```



#### d)

Uncorrected values

##### &ensp;i.
```{r}
M2.SSRu <- t(M2.yhat) %*% M2.yhat
M2.SSRu
```

```{r}
M2.SSE <- t(M2.ehat) %*% M2.ehat
M2.SSE
```


```{r}
M2.SSTu <- t(M2.y) %*% M2.y
M2.SSTu

# Are these equal?
all.equal(M2.SSTu, M2.SSRu + M2.SSE)
```

##### &ensp;ii.

Corrected values

```{r}
X1 <- matrix(rep(1, NROW(M2.Hat)))
H1 <- X1 %*% solve(t(X1)%*%X1) %*% t(X1)
M2.SSR <- t(M2.y) %*% (M2.Hat - H1) %*% M2.y
M2.SSR
```

```{r}
M2.SST <- M2.SSTu - t(M2.y) %*% H1 %*% M2.y
M2.SST

# Are these equal?
all.equal(M2.SST, M2.SSR + M2.SSE)
```

#### e)

```{r}
anova(M2)
```

First off, R makes it easy to see that only Race is a significant factor  (at the 90% level) after considering Smoking and Lwtkg.

This means we reject $H_0$: All group means are equal.


However, in the following table:

```{r}
summary(M2)$coefficients
```

we see that smoke is considered a significant factor. This might be an issue with the order of the F-tests, indicating a lack of independence among the partial F tests.

#### f)

##### &ensp;i.
```{r}
plot(bwgt$Race, M2$residuals, xlab='Race', ylab='residuals')
plot(bwgt$Smoke, M2$residuals, xlab='Smoke', ylab='residuals')
plot(bwgt$Lwtkg, M2$residuals, xlab='Lwtkg', ylab='residuals')
plot(M2$fitted.values, M2$residuals, xlab='Fitted', ylab='residuals')
```


Roughly same mean and equal variance in Race, Smoke.


##### &ensp;ii.

$H_0$: $\beta_1 = -1$
$H_1$: $\beta_1 \neq -1$

```{r}
# t = (B1 estimate - B1 under H0) / se(B1)
tstat <- (0.425788904  - (-1)) / 0.16689290
pvalue <- 2*pt(abs(tstat), df=df.residual(M2), lower.tail=F)
pvalue
```

There is very significant evidence to reject H0.

```{r}
alpha <- 0.05
tcrit <- qt(alpha/2, df=df.residual(M2), lower.tail=F)

tvec <- seq(length=101, from=-3, to=3)
dvec <- dt(tvec, df=df.residual(M2))
plot(tvec, dvec, type="l", xlab="t", ylab="P(t)",
     main=expression(t[15]*" Distribution for Smoke"))

abline(v=tstat, col="red")
tvec1 <- seq(length=31, from=tcrit, to=max(tvec))
dvec1 <- dt(tvec1, df=df.residual(M2))
polygon(  c(tvec1,rev(tvec1)), c(dvec1,rep(0,31)), col="grey")
polygon( -c(tvec1,rev(tvec1)), c(dvec1,rep(0,31)), col="grey")
```


### Question 2.

#### a)

```{r}
Observations <- c(7.2, 4.5, 6.6, 3.3, 11.2, 8.7)
Treatments <- as.factor(c(1, 1, 2, 2, 3, 3))
treatments <- data.frame(Treatments, Observations)

fit1 <- lm(Observations~Treatments, data=treatments)
anova(fit1)
```



#### b)

##### &ensp;i.
MSE = 4.0717

```{r}
q2.lambda = function(a1) {
  2 / 4.0717 * sum(
    sapply(
      c(a1, 0, -a1), 
      function(a) a^2
    
  ))
}
q2.powers <- function(alpha) {
  sapply(
    # alpha1 values
    X = c(1.0, 1.5, 2, 2.5, 3), 
    # Find power at the 95% quartile for this df
    FUN = function(a1) 
      pf(
        qf(1-alpha, df1=2, df2=3), 
        df1=2, 
        df2=3, 
        q2.lambda(a1)
      )
  )
}
q2.powers(0.05)
  
```

##### &ensp;ii.

```{r}
q2.powers(0.10)
```

